---
title: "Calculating inventory attributes using Forest Tools"
author: "Andrew Plowright"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating inventory attributes using Forest Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r global_options, include=FALSE, dpi =  300}
knitr::opts_knit$set(global.par = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include=FALSE}
library(ForestTools)
```

## Introduction

This short guide will demonstrate how to use the Forest Tools package to generate raster-based spatial statistics from a set of dominant treetops and then summarize the raster to a set of polygonal management units.

The analysis workflow is as follows:

1. Dominant treetops are detected from a canopy height model.
2. The average height of the 100 tallest trees is calculated for each 1 ha cell in a continuous grid. This metric will be referred to as _top height_.
3. The average top height is upscaled a second time to a set of polygonal management units (blocks).

The process for detecting dominant treetops (step 1) is explained in detail in the accompanying [Canopy analysis](treetopAnalysis.html) vignette. In this document, we will simply use an existing set of treetops.

Begin by loading the Forest Tools library and sample data. The sample data is derived from a photogrammetric survey of a 125 ha area in the Quesnel Timber Supply Area in British Columbia, Canada.

```{r, message = FALSE}
# Attach the 'ForestTools' and 'raster' libraries
library(ForestTools)
library(raster)

# Load sample canopy height model, treetops and block boundaries
data("quesnelCHM", "quesnelTrees", "quesnelBlocks")
```

View the canopy height model (CHM) and the polygonal management units (blocks) using the plot function.

```{r, fig.width = 4, fig.height = 2.51}
# Remove plot margins (optional)
par(mar = rep(0.5, 4))

# Plot CHM and blocks (extra optional arguments remove labels and tick marks from the plot)
plot(quesnelCHM, xlab = "", ylab = "", xaxt='n', yaxt = 'n')
plot(quesnelBlocks, add = TRUE, border =  "darkmagenta", lwd = 2)
```

## Compute top height for 1 hectare grid

Given the large number of treetops in the `quesnelTrees` dataset, visualizing it in its entirety is difficult. Instead, it is more practical to view a rasterized representation of the trees' attributes. This can be done using the `sp_summarise` function.

The first step is to define a function for computing the statistic that we're interested in. In this case, we want the average height of the tallest 100 trees. It is important, however, that our function accept a number of trees lower than 100 without returning an error---in this case, it should simply return the average height of all trees in the area. We should also consider the possibility that some trees will have NA (non-available) height values.

```{r}
# Create custom function for computing top height
topHgtFun <- function(x, ...) mean(tail(sort(x), 100))
```

In this example `x` would be a series of tree heights. The `sort` function will arrange them in ascending order (and also remove NA values), and the `tail` function will subset the last 100 values (i.e.: the highest values). Finally, the `mean` function will compute the average.

We will now apply the newly created `topHgtFun` to our `quesnelTrees` dataset using the `sp_summarise` function. We set the `grid` argument to 100 to generate a 100 m x 100 m (1 ha) grid. It is important to enter `topHgtFun` as a _named list_: "Top100" will be part of the name given to the product of the function. If other statistics are required, their corresponding functions can be added to this list.

```{r}
# Use sp_summarise to generate gridded statistics
sptStatRas <- sp_summarise(trees = quesnelTrees, variables = "height", grid = 100, statFuns = list(Top100 = topHgtFun))

# View information about the result
sptStatRas
```

The product of `sp_summarise` is a [RasterBrick](https://cran.r-project.org/package=raster/raster.pdf#page=36), i.e.: a multi-layered raster. It contains two layers: _TreeCount_ and _heightTop100_, the product of our custom function when applied to the _height_ attribute of the trees in each cell. We can subset a single layer from the RasterBrick using the `[[]]` operator

```{r, fig.width = 4, fig.height = 2.51}
# Subset top height raster
topHgtRas <- sptStatRas[["heightTop100"]]

# View top height on a 1 ha grid
plot(topHgtRas, xlab = "", ylab = "", xaxt='n', yaxt = 'n')
```

## Compute average cell height for polygonal areas

Once our raster-based statistic has been generated, there are a variety of ways summarizing the raster to a set of polygons. Note that this is different than calculating _top height_ for the polygons directly. In the following example, the `rasterize` and `zonal` functions from the [raster]( https://cran.r-project.org/package=raster) package are used.

```{r, fig.width = 4, fig.height = 2.51}
# Rasterize block boundaries
blockRas <- rasterize(quesnelBlocks, topHgtRas)

# View results
plot(blockRas, xlab = "", ylab = "", xaxt='n', yaxt = 'n')

# Use rasterized block boundaries to compute zonal statistics
zoneStat <- zonal(topHgtRas, blockRas, 'mean')
zoneStat
```

We can now simply attach the zonal statistics to our original polygonal areas.

```{r, fig.width = 4, fig.height = 2.51, message = FALSE}
# Create new 'topHeight' attribute from zonal statistics
quesnelBlocks[["topHeight"]] <- zoneStat[,"mean"]

# Plot result
library(rgeos)
colRamp <- colorRampPalette(c('lightgoldenrod1', 'tomato2'))(10)
polyCols <- colRamp[as.numeric(cut(quesnelBlocks[["topHeight"]],breaks = 10))]
plot(quesnelBlocks, col = polyCols, xlab = "", ylab = "", xaxt='n', yaxt = 'n')
text(gCentroid(quesnelBlocks, byid = TRUE), round(quesnelBlocks[["topHeight"]],2), font = 2)
```
